---
title: "experiments"
author: "Justin Lin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("out of sample algs.R")
library(ggplot2)
library(gridExtra)
library(rgl)
```

## Min Cost vs. Trosset vs. Bengio
The data is uniform on the cube [-5,5]^3 and the new point is chosen randomly in
the cube [-10,10]^3. The data was projected to two dimensions using the Bengio
method. This 2D embedding was used to calculate the Min Cost and Trosset points.

```{r Min Cost vs. Trosset vs. Bengio}
n = 20
data = matrix(runif(3*n, -5, 5), ncol = 3)
proj_data = Bengio_data(data, 2, K_mds)

new_pt = runif(3,-10,10)

a2 = vector(length = n)
for (i in 1:n) {
  a2[i] = dist_sq(new_pt, data[i,])
}

min_cost_pt = min_cost(proj_data, a2, range = 20)$par

Trosset_pt = Trosset(proj_data, compute_dist_sq_mat(proj_data), a2, range = 20)$par

Bengio_pt = Bengio(data, new_pt, 2, K_mds)

plot_data = data.frame(x = c(proj_data[,1], min_cost_pt[1], Trosset_pt[1], Bengio_pt[1]),
                       y = c(proj_data[,2], min_cost_pt[2], Trosset_pt[2], Bengio_pt[2]),
                       type = c(rep("Data", n), "Min Cost", "Trosset", "Bengio"))
ggplot(plot_data, aes(x = x, y = y, color = factor(type))) + geom_point()
```

## Relative Error: Min Cost vs. Trosset
The data was generated randomly on the square [-10,10]^2. The new point was 
randomly generated on the cube [-10,10]^2 x [0,20]. The square inter-point
distances of the embedded point were compared to the squared inter-point 
distances of the original point. Relative error means the absolute error were 
divided by the squared inter-point distances of the original point. The ten
simulations are in order based on the "Euclidean-ness" of the data.

```{r, relative error Min Cost vs. Trosset}
experiment = function(n) {
  proj_data = matrix(runif(2*n, -10, 10), ncol = 2)
  D2 = compute_dist_sq_mat(proj_data)
  
  rand_point = c(runif(2, -20, 20), runif(1, 0, 20))
  a2 = vector(length = n)
  for (i in 1:n) {
    a2[i] = dist_sq(c(proj_data[i,],0), rand_point)
  }
  
  new_D2 = rbind(cbind(D2,a2),c(a2,0))
  e = rep(1,n+1)
  P = diag(n+1) - e %*% t(e) / (n+1)
  B = -0.5 * P %*% new_D2 %*% P
  d = svd(B)$d
  emb = round((d[1] + d[2]) / sum(d), digits = 5)
  
  Trosset_pt = Trosset(proj_data, D2, a2, range = 20)
  Trosset_error = rel_error(Trosset_pt$par, proj_data, a2)
  
  min_cost_pt = min_cost(proj_data, a2, range = 20)
  min_cost_error = rel_error(min_cost_pt$par, proj_data, a2)
  
  data.frame(Ddata_sq = a2, Trosset_error, min_cost_error, emb = rep(emb,n))
}

get_data = function(n, runs) {
  data = experiment(n)
  
  for (i in 2:runs) {
    data = rbind(data, experiment(n))
  }
  data
}

data = get_data(60,10)

p = ggplot(data, aes(x = Ddata_sq, y = Trosset_error, col = "Trosset")) + 
  geom_point() +
  geom_smooth(se = FALSE) + 
  geom_point(data = data, aes(x = Ddata_sq, y = min_cost_error, col = "Min Cost")) +
  geom_smooth(data = data, aes(x = Ddata_sq, y = min_cost_error, col = "Min Cost"), se = FALSE)
p + facet_grid(rows = vars(emb), scale = "free_y")
```


## Reproducing inter-point distances with error: Min Cost vs. Bengio
The data was uniform in the solid sphere with radius 5. The new points were
distances of 2, 5, and 10 away from the origin. The Gaussian error had standard
deviation of 1, 2, 5, and 10 in all three coordinates. The Min Cost point was
calculated using the 2D Bengio embedding. The mean relative errors in inter-
point distances was calculated for both methods in each trial.

```{r, reproducing inter-point distances with error Min Cost vs. Bengio}
get_rel_errors = function(data, proj_data, base_pt, error_sd) {
  error_pt = base_pt + rnorm(3, 0, error_sd)
  Bengio_pt = Bengio(data, error_pt, 2, K_mds)
  
  n = length(data[,1])
  a2 = vector(length = n)
  for (i in 1:n) {
    a2[i] = dist_sq(error_pt,data[i,])
  }
  min_cost_pt = min_cost(proj_data, a2, range = 15, iter = +1e06)$par
  
  base_pt_dists = vector(length = n)
  Bengio_dists = vector(length = n)
  min_cost_dists = vector(length = n)
  
  for (i in 1:n) {
    base_pt_dists[i] = dist_sq(base_pt, data[i,])
    Bengio_dists[i] = dist_sq(Bengio_pt, proj_data[i,])
    min_cost_dists[i] = dist_sq(min_cost_pt, proj_data[i,])
  }
  
  cbind(abs(base_pt_dists - min_cost_dists)/base_pt_dists, 
        abs(base_pt_dists - Bengio_dists)/base_pt_dists)
}

sim = function(n = 20, b = 100, new_pt, sd) {
  theta = runif(n,0,2*pi)
  phi = runif(n, 0, pi)
  r = 5*(runif(n,0,1))^(1/3)
  
  x = r*sin(phi)*cos(theta)
  y = r*sin(phi)*sin(theta)
  z = r*cos(phi)
  
  data = matrix(c(x,y,z), ncol = 3)
  proj_data = Bengio_data(data, 2, K_mds)
  
  mean_errors = matrix(nrow = b, ncol = 2)
  for (i in 1:b) {
    errors = get_rel_errors(data, proj_data, new_pt, sd)
    mean_errors[i,] = colMeans(errors)
  }
  
  mean_errors
}

build_box = function(sds, base_pt, data_size = 20) {
  n = length(sds)
  mat = matrix(nrow = n*100, ncol = 3)
  
  for(i in 1:n) {
    mat[((i-1)*100+1):(i*100),1] = rep(sds[i],100)
    mat[((i-1)*100+1):(i*100),2:3] = sim(n = data_size, new_pt = base_pt, sd = sds[i])
  }
  table = data.frame(sd = c(mat[,1],mat[,1]),
                     mean_rel_error = c(mat[,2],mat[,3]),
                     type = c(rep("Min Cost", n*100), rep("Bengio", n*100)))
  
  ggplot(table, aes(x = factor(sd), y = mean_rel_error, color = type)) + geom_boxplot()
}

p1 = build_box(c(1,2,5,10), c(2,0,0))
p2 = build_box(c(1,2,5,10), c(5,0,0))
p3 = build_box(c(1,2,5,10), c(10,0,0))
grid.arrange(p1,p2,p3,ncol=1)
```


## Reproducing inter-point distances on the sphere with error: Min Cost vs. Bengio
The data was uniform on the upper hemisphere of radius 5 with geodesic 
distances. The Min Cost point was calculated using the 2D Bengio embedding. The
mean relative errors in inter-point distances was calculated for both methods in 
each trial.

```{r, reproducing inter-point distances on the sphere}
get_rel_errors = function(spherical_data, proj_data, base_pt, error_sd) {
  n = length(spherical_data[,1])
  
  error_theta = rnorm(1, 0, error_sd)
  error_phi = rnorm(1, 0, error_sd/4)
  error_pt = base_pt + c(error_theta, error_phi)
  error_pt_cart = c(5*sin(error_pt[2]*cos(error_pt[1])), 
                    5*sin(error_pt[2]*sin(error_pt[1])),
                    5*cos(error_pt[2]))
  
  x = 5*sin(spherical_data[,2])*cos(spherical_data[,1])
  y = 5*sin(spherical_data[,2])*sin(spherical_data[,1])
  z = 5*cos(spherical_data[,2])
  
  cartesian_data = cbind(x,y,z)
  
  Bengio_error_pt = Bengio(cartesian_data, error_pt_cart, 2, K_mds)
  
  a2_error_pt = vector(length = n)
  for (j in 1:n) {
    a2_error_pt[j] = spherical_dist_sq(5, error_pt, spherical_data[j,])^2
  }
  min_cost_error_pt = min_cost(proj_data, a2_error_pt, range = 5, iter = 1e+06)$par
  
  base_pt_dists = vector(length = n)
  Bengio_dists = vector(length = n)
  min_cost_dists = vector(length = n)
  
  for (i in 1:n) {
    base_pt_dists[i] = spherical_dist_sq(5, base_pt, spherical_data[i,])
    Bengio_dists[i] = dist_sq(Bengio_error_pt, proj_data[i,])
    min_cost_dists[i] = dist_sq(min_cost_error_pt, proj_data[i,])
  }
  
  cbind(abs(base_pt_dists - min_cost_dists)/base_pt_dists, 
        abs(base_pt_dists - Bengio_dists)/base_pt_dists)
}

sim = function(n = 20, b = 100, new_pt, sd) {
  theta = runif(n,0,2*pi)
  phi = runif(n, 0, pi)
  r = 5*(runif(n,0,1))^(1/3)
  
  spherical_data = cbind(theta,phi)
  
  x = r*sin(phi)*cos(theta)
  y = r*sin(phi)*sin(theta)
  z = r*cos(phi)
  
  cartesian_data = matrix(c(x,y,z), ncol = 3)
  proj_data = Bengio_data(cartesian_data, 2, K_mds)
  
  mean_errors = matrix(nrow = b, ncol = 2)
  for (i in 1:b) {
    errors = get_rel_errors(spherical_data, proj_data, new_pt, sd)
    mean_errors[i,] = colMeans(errors)
  }
  
  mean_errors
}

build_box = function(sds, base_pt, data_size = 20) {
  n = length(sds)
  mat = matrix(nrow = n*100, ncol = 3)
  
  for(i in 1:n) {
    mat[((i-1)*100+1):(i*100),1] = rep(sds[i],100)
    mat[((i-1)*100+1):(i*100),2:3] = sim(n = data_size, new_pt = base_pt, sd = sds[i])
  }
  table = data.frame(sd = c(mat[,1],mat[,1]),
                     mean_rel_error = c(mat[,2],mat[,3]),
                     type = c(rep("Min Cost", n*100), rep("Bengio", n*100)))
  
  ggplot(table, aes(x = factor(sd), y = mean_rel_error, color = type)) + geom_boxplot()
}

p1 = build_box(sds = c(pi/16,pi/8,pi/4), base_pt = c(0,pi/8))
p2 = build_box(sds = c(pi/16,pi/8,pi/4), base_pt = c(0,pi/4))
p3 = build_box(sds = c(pi/16,pi/8,pi/4), base_pt = c(0,pi/2))
grid.arrange(p1,p2,p3,ncol = 1)
```


## Projecting a point on the sphere with error: Min Cost vs. Bengio
The data was uniform on the upper hemisphere of radius 5 with geodesic 
distances. The Min Cost point was calculated using the 2D Bengio embedding. The
projected data, Bengio embedding of the base point, Bengio embedding of the
noised point, and Min Cost embedding of the noised point are plotted.

```{r, error point on sphere}
sim = function(n, new_pt, error_sd) {
  theta = runif(n,0,2*pi)
  phi = runif(n, 0, pi)
  r = 5*(runif(n,0,1))^(1/3)
  
  spherical_data = cbind(theta,phi)
  
  x = r*sin(phi)*cos(theta)
  y = r*sin(phi)*sin(theta)
  z = r*cos(phi)
  
  cartesian_data = matrix(c(x,y,z), ncol = 3)
  proj_data = Bengio_data(cartesian_data, 2, K_mds)
  
  new_pt_cart = c(5*sin(new_pt[2]*cos(new_pt[1])), 
                  5*sin(new_pt[2]*sin(new_pt[1])),
                  5*cos(new_pt[2]))
  
  Bengio_base = Bengio(cartesian_data, new_pt_cart, 2, K_mds)
  
  error_theta = rnorm(1, 0, error_sd)
  error_phi = rnorm(1, 0, error_sd/4)
  error_pt = new_pt + c(error_theta, error_phi)
  error_pt_cart = c(5*sin(error_pt[2]*cos(error_pt[1])), 
                    5*sin(error_pt[2]*sin(error_pt[1])),
                    5*cos(error_pt[2]))
  
  Bengio_pt = Bengio(cartesian_data, error_pt_cart, 2, K_mds)
  
  a2 = vector(length = n)
  for (i in 1:n) {
    a2[i] = spherical_dist_sq(5, error_pt, spherical_data[1,])^2
  }
  min_cost_pt = min_cost(proj_data, a2, range = 5, iter = 1e+07)$par
  
  data = data.frame(x = c(proj_data[,1], Bengio_base[1], min_cost_pt[1], Bengio_pt[1]), 
                    y = c(proj_data[,2], Bengio_base[2], min_cost_pt[2], Bengio_pt[2]),
                    type = c(rep("Data", n), "Bengio Base", "Min Cost", "Bengio"))
  ggplot(data, aes(x = x, y = y, color = factor(type))) + geom_point()
}

sim(60, new_pt = c(0, pi/4), error_sd = pi/4)

```

